Formal Analysis of EDHOC Key Establishment for Constrained IoT Devices

### 2022-07-21 paper.tex

%-------------------------------------------------------------------------- sec
\section{\uppercase{Trusted Execution Environments}}
\label{sec:TEE}
The main purpose of Trusted Execution Environments (TEE) is to prevent an
adversary with physical access to a device from accessing the computational
circuits.
%
They come in various forms of differing complexity.
%
Some, like ARM Trust Zone and Intel SGX are general purpose execution
environments, which are flexibly programmable.
%
Other, like the Universal Subscriber Identity Modules (USIM) used for
authentication to 3GPP mobile networks, have application specific interfaces for
authentication and key agreement protocols etc.
%
A key aspect is how much of the application is placed in the TEE and how much is
outside.
%

For larger devices that include general purpose processors with Trust Zone or
SGX, entire \mEdhoc{} and \mOscore{} may reside inside the TEE.
%
For constrained IoT devices in the lower end of the scale, a TEE may have to be
implemented using a special-purpose integrated circuit.
%
In the latter case, it may be beneficial to follow a minimalistic approach, and
contain the long-term key and only
the operations that need access to it in the TEE for cost reasons.
%

It may at first appear more secure to implement as much as possible inside the
TEE, but there is a security trade-off: the more code inside the TEE,
the higher the risk of implementation errors in the security-critical code.
%
Because the security critical code runs in the area where the long-term keys
reside, an implementation error here risks leaking information of the key to the
adversary.
%
From this perspective, it may be beneficial to follow the minimalistic approach
even when having access to Trust Zone or SGX.
%

A slightly more secure division of functionality is to also keep the
session key inside
the TEE and extend the interface to accept messages and return the
(en/de)crypted counter part.
%
That is, the interface exposes \mAead{} functions in the interface.
%

%-------------------------------------------------------------------------- sub
\subsection{Practical Aspects and Use Case}
\label{sec:TEE:useCase}
Authentication of a specific IoT device assumes that this devices is the only
entity with access to the long-term key associated with the corresponding
identity.
%
Since IoT devices may be accessible to adversaries, e.g., an insider cloning a
key card, the long-term keys must be appropriately protected.
%
A state of the art approach is to use a Trusted Execution Environment (TEE),
which holds the key and provides an API for operations using the key.
%
This is the approach taken by the
Trust-Zone based \mMuEdhoc{}~\cite{DBLP:conf/codaspy/Hristozov0XFLS21}
implementation for example.
%
Typical operations include signatures using the long-term private key of a
party.
%

%-------------------------------------------------------------------------- sub
\subsection{System and Adversary Model Extensions Supporting TEE}
\label{sec:TEE:advModel}
%
We extend the adversary model of~\cite{Norr21} by allowing the adversary
access to a TEE interface on a party of their choosing before and after the
\mEdhoc{} session.
%
This allows us to distinguish between a full compromise and a weak one, using
the terminology of~\cite{DBLP:conf/csfw/Cohn-GordonCG16}.
%
A weak compromise is when the adversary have temporary access to the TEE.
%
Seen in the framework of~\cite{DBLP:conf/icics/XuZRWTZ20}, we add adversary
capabilities corresponding to a server adversary, i.e., an adversary that
compromises a party, learns its ephemeral keys and may temporarily access its
TEE, but does not learn the party's long-term key.
%
The model of~\cite{Norr21} only captures server adversary capabilities without
access to the TEE, via the leakage of ephemeral keys, and local adversary
capabilities, or more specifically, a secure-world adversary with read-write
access to the
secure world where the long-term keys reside via the leakage of long-term keys.
%
Our extension allows analysis of so-called weak post-compromise security
(PCS)~\cite{DBLP:conf/csfw/Cohn-GordonCG16}.
%

We split the \mEdhoc{} functionality as follows.
%
The TEE contains the long-term key and allows the non-TEE parts of the
application to perform the operations on it via its interface.
%
More precisely, parties using the \mSig{} authentication method use a TEE with
an interface which accepts a message and returns the signature of that message
using the party's private long-term key.
%
A party $U$ using the \mStat{} authentication method use a TEE with an interface
accepting a point $P$ on the curve and returning $\mDH(P,\ \mGu{})$.
%
\knote{It should accept any value. Adversary may not stick to elements of the
    form $g^x$. Or we explicitly state that the TEE checks that the received
input lies on the curve (but this may not be sufficient: are the curves cyclic
groups? Need to check).}
%
The non-TEE parts implement \mEdhoc{} using this interface.
%
This interface requires the least functionality from the TEE, reducing the TEE's
complexity and possibly cost of its implementation.
%
The functional split is suitable even for the situation where a
constrained device has implemented only the storage of the long-term key in a
special purpose circuit with minimal processing functionality.
%
Since \mEdhoc{} focuses on constrained IoT devices, it seems appropriate to
cater for this setting.
%

%-------------------------------------------------------------------------- sub
\subsection{Extended Formalism and Security Properties}
\label{sec:TEE:fmAndProps}
Formally, we model the TEE interface by adding two new rewrite rules:
%
\begin{small}
\begin{verbatim}
rule forge_SIG:
   [!LTK_SIG($A, ~ltk), In(xx)] --[TEE($A)]-> [Out(sign(xx, ~ltk))]

rule exp_STAT:
   [!LTK_STAT($A, ~ltk), In('g'^~xx)] --[TEE($A)]-> [Out(('g'^~xx)^~ltk)]
\end{verbatim}
\end{small}
%
These rules allow the adversary to obtain terms representing signatures on a
value of their choice to forge signatures (\verb|forge_SIG|), or to obtain terms
representing a curve point of their choice raised to the power of the
long-term key (\verb|exp_STAT|).
%

Because it is a trivial attack when the adversary access these rules with values
from the test session, we must disqualify those rule applications.
%
We do so by creating an action fact \verb|TEE($A)|, where \verb|$A| is the
identity corresponding to the private key used, and then augmenting the
properties with a condition that no such action fact exists from the start of
the protocol execution and its end.
%
Care must be taken when specifying the start and the end.
%
Specifically, the start and end of the protocol run must be viewed w.r.t. the
role.
%
For example, the injective agreement property for the initiator in
the \mSigSig{} method requires that the adversary does not have access to the
TEE of the
responder from the time of transmission of the first message until the second
message is received by the initiator.
%
Because reception of one message and transmission of the next one at a party
is an atomic operation, these timepoints represent the start and end of the
protocol run from the perspective of the initiator.
%

We did not manage to get Tamarin to terminate for all lemmas and all methods
with the added TEE extension.
%
When the responder uses the \mStat{} authentication method Tamarin did not
complete the in a reasonable time for the authentication lemmas.
%
To combat this, we simplified the modeling of XOR encryption of the second
message for these lemmas when the responder uses \mStat{} authentication.
%
More precisely, the second message contains the two values \verb|R| (the
responder's identity) and
\verb|authR| (the authentication information), and we generally model the
XOR encryption by xoring each of them with their own key-stream terms.
%
For the problematic lemmas however, we resorted to xoring the entire tuple
\verb|<R, authR>| with a single key-stream term.
%
Given that no attacks were identified using our original modeling, this
simplification is intuitively not a severe restriction.
%
However, it is possible that we now miss attacks on these methods
combining properties of the two-term XOR encryption with TEE capabilities.
%

@@

\subsubsection{Non-Repudiation:}
Access control mechanisms can vary depending on what the underlying application is.
%
A nuclear power-plant might need to keep track of who enters and leaves, but a coffee machine that logs every user and their coffee preference might lead to a privacy concern.
%
This thought experiment allowed us to perceive that the \mEdhoc{} \mSpec{} did not even consider non-repudiation.
%
When we pointed this out to the authors of the \mSpec, they recognized this concern, and added a discussion about how the various \mEdhoc{} methods satisfy (or not) non-repudiation.

\subsubsection{Unintended Peer Authentication: }
In Section~3.2 in the \mSpec{}, the authors state that parties executing \mEdhoc{} must be configured such that they only run \mEdhoc{} with a restricted (according to some reasonable policy) set of peers.
%
However, \mIdcredr{} in the second message is sent in the clear, and can therefore lead to some masquerade attacks by the adversary.
%

@@

\subsubsection{Denial of Service: }
Denial of Service (DoS) attacks for IoT protocols are very well-studied.
%
But DoS is not well documented for \mEdhoc{} in particular, since it delegates DoS handling to the application.
%
The \mSpec{} does not include any specific measures for countering such attacks.

>>>>

@@

\subsubsection{Non-Repudiation.}
Access control mechanisms can vary depending on what the underlying application is.
%
A nuclear power-plant might need to keep track of who enters and leaves, but a coffee machine that logs every user and their coffee preference might lead to a privacy concern.
%
This thought experiment allowed us to perceive that the \mEdhoc{} \mSpec{} did not even consider non-repudiation.
%
When we pointed this out to the authors of the \mSpec, they recognized this concern, and added a discussion about how the various \mEdhoc{} methods satisfy (or not) non-repudiation.

\subsubsection{Unintended Peer Authentication.}
\label{sec:unintendedPeerAuth}
In Section~3.2 in the \mSpec{}, the authors state that parties executing \mEdhoc{} must be configured such that they only run \mEdhoc{} with a restricted (according to some reasonable policy) set of peers.
%
However, \mIdcredr{} in the second message is sent in the clear, and can therefore lead to some masquerade attacks by the adversary.
%

@@

\subsubsection{Denial of Service.}
Denial of Service (DoS) attacks for IoT protocols are very well-studied.
%
But DoS is not well documented for \mEdhoc{} in particular, since it delegates DoS handling to the application.
%
The \mSpec{} does not include any specific measures for countering such attacks.

### 2022-07-21 protocol.tex

%------------------------------------------------------------------------- sec
\section{Implementation Aspects and Key Protection}
\label{sec:TEE}
Adversaries may have physical access to implementations, this is perhaps an
even higher risk in an unsupervised IoT setting.
%
Implementations can make use of Trusted Execution Environments (TEE) hardened
against physical attacks to mitigate adversaries with physical access to the
cryptographic keys.
%

>>>>

%------------------------------------------------------------------------- sec
\section{Implementation Aspects and Key Protection}
\label{sec:TEE}
Authentication of a specific IoT device assumes that this devices is the only
entity with access to the long-term key associated with the corresponding
identity.
%
Since IoT devices may be accessible to adversaries, e.g., an insider cloning a
key card, the long-term keys must be appropriately protected.
%
A state of the art approach is to use a Trusted Execution Environment (TEE),
which holds the key and provides an API for operations using the key.
%
This is the approach taken by the
Trust-Zone based \mMuEdhoc{}~\cite{DBLP:conf/codaspy/Hristozov0XFLS21}
implementation for example.
%
Typical operations include signatures using the long-term private key of a
party.
%
